Declarative Probabilistic Logic Programming\\in Discrete-Continuous Domains

### 2024-09-10 202409_aij_dc_problog_aij_second/rebuttal.tex

\section*{ \Huge Rebuttal}
Firstly, we would like to thank all reviewers for taking their time and giving such detailed comments. We understand that our manuscript is rather long and that reviewing it takes considerable effort. We would like to stress that we found the reviews detailed and very helpful in improving the manuscript. Please find below our detailed rebuttal.

\section*{General Comments}

In the following sections we address the points raised by the reviewers one by one. We highlight in \fixedrebuttal{green} where we agree with the reviewers' opinion and where we made adequate changes in the manuscript. We highlight in \commentrebuttal{blue} parts where we do not necessarily agree with a certain opinion and give our justification. Within the paper there are passages that are highlighted in \new{purple}, which indicates new material not present in the previous version of the paper.

We would first like to highlight the three major changes made to the manuscript compared to the initial submission:
\begin{enumerate}
    \item As asked by Reviewer 1 we have added a proof that the $sigma$-algbera of the probability measure $P_F$ is indeed (as suggested by the reviewer) a cylinder measure. This let (in our opinion) to a more streamlined exposition of the semantics. In the draft of the paper this manifests itself by the three new Propositiona \ref{prop:omegaf} \ref{prop:pfsigma} and \ref{prop:pf}.
    \item We have relegated large parts of the syntactic sugar discussion to the appendix (as suggested by Reviewer 3). We hope this increases readability and accessibility of the manuscript. Specifically, we only introduce the syntactic sugar by example and discuss the technical details in the appendix.
    \item We have added experimental evidence (\cf Section~\ref{sec:experimental}) that our proposed algorithm for the discrete-continuous domain ( (Symbolic IALW)) results in improvements over state-of-the-art approaches.
\end{enumerate}

\section{Reviewer 1}

\subsection{Sigma Algebra}
Overall I believe the paper deserves publication but there is one point that should be clarified: in the proof of
proposition 3.20 that a valid DF-PLP program induces a unique probability distribution over Herbrand interpretations,
eq (C.3), it is not obvious that the set $\{F_\omega|M_\omega(V)\models \mu_1^{b_1} \wedge ... \wedge \mu_n^{b_n}\}$ is
measurable according to $P_F$. In proposition 3.15 the authors proved that $P_F$ exists and is unique but provide no
information on the $\sigma$-algebra of the probability measure. I think the algebra of $P_F$ is the sigma algebra of
cylinder sets of value assignments to $F$ but in this case this is not the powerset so it is not evident that
the set above belongs to the sigma algebra of $P_F$. This should be proved.

\fixedrebuttal{We agree with the reviewer that this was missing. We took the opportunity and restructure the construction of the probability space (using the new Definition~\ref{def:distDB}), which then led to a more explicit construction of the meaure $\probabilitymeasure_\comparisonfacts$. See also our first comment in the general comments above. }

\subsection{Minor comments}

\begin{enumerate}
    \item Example 5.3 and following: why not using a program clause instead of the mathematical definitions of $\nu_1$ and$ \nu_2$?
    \\
    \commentrebuttal{We strictly restricted the clause notation to writing explcit programs. We used math notation everywhere else. This is due to the fact that using clause notation makes it occasionally hard to read formal expression, especially when we make use of indexing.}

    \item Example 5.4: $\eta=1-> \eta=\top$
    \\
    \fixedrebuttal{Fixed.}
    \item Note 3: for a program to be range restricted, all the variables in the head should occur in POSITIVE literals in the body.
    \\
    \fixedrebuttal{Fixed.}
    \item Line 802: Definition 4.14 repeated
    \\
    \fixedrebuttal{Fixed.}
    \item Eq 7.3: I believe $N$ should be $|S|$
    \\
    \fixedrebuttal{Fixed.}
    \item Eq 7.10 $(s,m)->(t,m)$
    \\
    \fixedrebuttal{Fixed.}
    \item line 920 $n>m$ -> n
    \\
    \fixedrebuttal{Fixed.}
    \item line 966 $\phi$->$\varphi$
    \\
    \fixedrebuttal{Fixed.}
    \item line 999: this seems to be removed
    \\
    \fixedrebuttal{Fixed.}
    \item note 5: "(otherwise case in Definition 7.12)"; Def 7.12 does not speak of circuits, why is it mentioned here?
    \\
    \fixedrebuttal{Fixed, Indeed we referenced the wrong definition. We changed the footnote to referencing Definition~\ref{def:sample_labeling_function}.}
    \item Algorithm 7.19: it should return a probability but it actually returns an infinitesimal number
    \\
    \fixedrebuttal{Fixed. We also slightly adapted the text describing the final step of the algorithm (just above Algorithm 7.19.)}
    \item line 1000: computeS
    \\
    \fixedrebuttal{Fixed.}
    \item Figure 7.4: what is the meaning of 1 in \texttt{size\_1(1)} and \texttt{size\_0(1)}? the part in parenthesis does not seem necessary
    \\
    \fixedrebuttal{Fixed. Indeed the part in the parenthesis is not necessary. This was a remnant from a previous version. We removed the parenthesis and its argument.}
    \item Page 44:
    $Eval(5)=\alpha_IALW(1=1)$->$\alpha_{IALW}(size_0=0.4)$
    \\
    \fixedrebuttal{Fixed.}
    \item line 1029 DC-ProbLog->ProbLog
    \\
    \fixedrebuttal{Fixed.}
    \item page 46 Eval formulas: some formulas have equalities with random variables, other equalities with constants

    $Eval(4)=..\alpha_{SIALW}(1=1)$->$Eval(4)=..\alpha_{SIALW}(m=1)$

     $Eval(2)=..\alpha_{SIALW}(0.3)$->$Eval(2)=..\alpha_{SIALW}(M=1)$

     $Eval(5)=\alpha_{IALW}(1=1)$->$\alpha_{IALW}(size_0=0.4)$
     \\
     \fixedrebuttal{Fixed.}
     \item line 1069 inference task repeated
     \\
     \fixedrebuttal{Fixed.}
    \item line 1075 remove "views"
    \\
    \fixedrebuttal{Fixed.}
    \item eq C.3 $F_\omega->F_\omega(V)$
    \\
    \fixedrebuttal{Fixed.}
    \item eq F.11 in the numberator $|S|$ is missing from the limit, in the denominator $N$ should be replaced by $|S|$
    \\
    \fixedrebuttal{Fixed.}
\end{enumerate}

\section{Reviewer 2}

\subsection{Weaknesses}

\begin{enumerate}
    \item The contribution of this work is more conceptual with a focus on extending Problog. It would be much better and more complete if the paper provides concrete evidence of how the extension could benefit real-life applications, and, especially how DC-ProbLog with IALW performs in comparison with related approaches. For example, whether the proposed language can be applied to perform reasoning of mixed variables at a large scale such as images and their labels. There are several attempts with neurosymbolic approach which deserve to be discussed in this paper.

    \fixedrebuttal{We have added an experimental section demonstrating the benefits of the proposed infernce algorithm (\cf Section~\ref{sec:experimental})}

    \fixedrebuttal{In the context of neurosymbolic AI we have already deployed our semantics by developing the neurosymbolic programming language DeepSeaProblog \citep{desmet2023neural}. We discuss this now in the related work section. Note that without the work performed in this paper, developing DeepSeaProblog would have constituted a rather intricate affair.}

    The paper should clarify the complexity of the approach, e.g. whether it is scalable in terms of presentation (space complexity) and inference (time complexity). For example, in example 2.2, if the depth of the relation children-parents-ancestors increases the number of distributional facts would grow exponentially.

    \commentrebuttal{Indeed in Example 2.2, the number of distributional facts would grow exponentially. Hinting at the computational hardness of the problem. Note, however, that here we are intersted in defining the semantics of a Turing complete probabilistic programming language. Meaning that performing inference in \dcproblogsty is actually an undecidable problem. While we do not state this explicitly this follows immediately from \dcproblogsty being Turing-complete.

    In practice, however, we are concerned with programs that will become ground eventually. In this respect we already state explicitly that computing algebraic model counts is \#P-hard (\cf Section~\ref{sec:ALWviaKC}).

    With these two points we believe we have sufficiently delineated the complexity of inference in \dcproblogsty.
    }

    \item The paper's presentation needs to be significantly improved in order to make it more easily comprehensible for a wider range of readers.

    \commentrebuttal{Following the suggestion from Reviewer 3 concerning the syntactic sugar we moved a large part from the main body of the manuscript to the appendix. We hope this improves readability.}

    The numbering is not well-formatted, it is mixed between examples, definitions, propositions, theorems, etc.

    \commentrebuttal{The AIJ author guide does not provide any details on how to number environments (e.g, propositions and theorems)\footnote{\url{https://www.sciencedirect.com/journal/artificial-intelligence/publish/guide-for-authors}}. In our opinion having different environments sharing the same numbering helps to navigate the paper as everything is in order. We would like to note that this is not uncommon. For instance, the template for the "International Conference on Machine Learning" follows a similar convention. }

    Some definitions are not well-written. For example Definition 4.14 is more like a construction/procedure. There are two definitions (3.7 and 4.17) for Parent and Ancestor. Please check other definitions as well.

    \fixedrebuttal{We renamed Definition~\ref{def:df_ancestor} from (Parent, Ancestors) to (Parents and Ancestors of Random Variables). Furthermore, we renamed Definition 4.17 (now Definition~\ref{def:parentancestor2}) from (Parent, Ancestors) to (Parents and Ancestors of Random Terms)}

    \commentrebuttal{While we agree that Definition 4.14 (now Definitino~\ref{def:elim-ad}) and Definition 4.25 (now Definition~\ref{def:adfree-to-core}) have a procedural character we think this is acceptable. Especially, as they define rewrite rules to be applied to a program.}

    There are too many definitions.

    \commentrebuttal{We hope that moving parts of the syntactic sugar discussion to the appendix sufficiently reduced the number of definitions.}

    The paper may be difficult for those who do not have a deep background in probabilistic logic programming. It is a good idea to have the "A Panoramic Overview" section early. However, it goes straight to the CD-Problog concept without laying the ground for readers to understand the basis of probabilistic logic programming.

    \fixedrebuttal{Fixed: we now give an introductory example in Section~\ref{sec:introduction} (Example~\ref{example:intro}).}

\end{enumerate}

\subsection{Other comments}

\begin{enumerate}
    \item Abstract: DC-ProbLog --> DC has not been defined yet.
    \\
    % \commentrebuttal{\dcproblogsty is the name of the language and therefore a proper name. We do not believe that it necessistates stating in the abstract that this stands for  "distributional clauses probabilistic Prolog", where "Prolog" stands for the French expression "programmation en logique".}
    \fixedrebuttal{Fixed.}
    \item 5-10: can be used represent --> to represent
    \\
    \fixedrebuttal{Fixed.}
    \item 55-60: such e.g --> such as
    \\
    \fixedrebuttal{Fixed.}
    \item 120-125: distributional facts in disguise --> please make the statement more formal
    \\
    \fixedrebuttal{Fixed. We now say: Note how probabilistic facts are actually syntactic sugar for distributional facts.}
    \item 135-14: DF-PLP: first time introduced, please provide the detailed name.
    \\
    \fixedrebuttal{Fixed.}
    \item Example 4.26: T2a and T2b are only mentioned once, the purpose of these notations should be well explained.
    \\
    \fixedrebuttal{Fixed. This was indeed a remnant from an earlier draft and the T2a and T2b are referring to the rules in Definition~\ref{def:adfree-to-core}. We now refer to these rules as CR1 and CR2. Note this part of the manuscript moved from the main body to the appendix (Line ~\ref{line:comment_r2_t2a_rule}).}
    \item 555-560: comparison literals --> check the grammar
    \\
    \commentrebuttal{We believe the grammar is correct. Note this part has moved to the Appendix in Line \lineref{line:comment_grammar_r2}}
    \item 555-560: maintains measurability of the latter --> the measurability
    \\
    \fixedrebuttal{Fixed.}
    \item before 1060: lsited bellow --> listed
    \\
    \fixedrebuttal{Fixed.}
\end{enumerate}

\section{Reviewer 3}

\subsection{Weaknesses}

\begin{enumerate}
    \setcounter{enumi}{0}

\item The proposal seems very similar to an approach called Bayesian Logic
Program, but with a different syntax. The semantics look very similar, and
the basic results are inherited from that work. The authors should clarify
the similarities and differences between their approach and Bayesian Logic
Programs, and provide sufficient reasons for introducing a new language.
This comparison, as far as I can see is not included in the related work
section at the end of the paper.

\fixedrebuttal{We have added a discussion on the relationship to Bayesian logic programs to the related work section (Section~\ref{sec:blp}). Following the suggestion from Reviewer 1 we reworked the theory in Section~\ref{sec:semantics}. We now substantially generaliize BLPs and do not inherit any of the results from \Citet{kersting2000bayesian} anymore. We discuss the differences in Section~\ref{sec:blp}.}

\item The most novel (and interesting) part of the paper concerns the infinites-
imal semantics, which comes very late in the paper. I have some doubts
about the formalisation. If I’m correct this could be a problem. (see
comments below)

\commentrebuttal{we discuss the reviewer's doubt in more detail in the technical comments section below. While pointing out some minor issues, we were able to readily resolve them.}

\item I have found the part on “syntactic sugar” not very relevant. It’s ok to
put it in a manual. It does not introduce new important concepts, and it
distracts the reader from the main message of the paper.

\fixedrebuttal{We relegated a large portion from the "syntactic sugar" discussion to the Appendix. In the main body of the paper we only discuss the syntax, while studying the semantics and specific details in the appendix.}

\item No experimental evaluation is provided.

\fixedrebuttal{We have now included an experimental comparison (Section~\ref{sec:experimental}). Furthermore, we would like to point out the experiments  we have performed in a(n) (already published) follow-up paper in the field of neurosymbolic AI \citep{desmet2023neural}.}

\end{enumerate}

\subsection{Readability}
\begin{enumerate}
    \setcounter{enumi}{7}

\item The paper is not easy to read. I believe it is not well organised. The most
important part is at the end of the paper. I believe the paper should concentrate mostly on basic primitives and leave the rewriting of ”syntactic
sugar primitives” later.

\fixedrebuttal{as mentioned above we have relegated large parts of the syntactic sugar discussion to the appendix.}

\item the paper is too long. There are many examples, even for elementary
concepts. The reader gets lost. I think this style is good for a manual,
not a scientific paper.

\commentrebuttal{removing the syntactic sugar discussion from the main body has significantly reduced the paper's length. Concerning the examples, we refer the reader to our comment on the next point (\cf AIJ author guidelines)}

\item I have found the panoramic overview not very interesting. It’s good to have
it in a manual but not in a scientific article. A more compact example
would do the job.

\commentrebuttal{It seems that this is a question of taste. Reviewer 1 and 2 seem to have a different opinion. Furtheermore, AIJ demands from papers to be accessible to a wider audience. Quoting from the AIJ author guidelines\footnote{https://www.sciencedirect.com/journal/artificial-intelligence/publish/guide-for-authors}: "Papers that are heavily mathematical in content are welcome but should include a less technical high-level motivation and introduction that is accessible to a wide audience and explanatory commentary throughout the paper." We believe the overview section and the examples throughout the paper have this function.}

\item The notation sometimes is heavy, I think that some simplifications are
possible.

\commentrebuttal{
We put considerable work into keeping the notation as light as possible yet consistent throughout the paper and also precise. The latter two affect of course, occasionally, readability. However, if there are concrete suggestions on how to improve the notation we would eagerly incorporate them into the manuscript.
} 
\end{enumerate}

\subsection{Technical Issues}

\begin{enumerate}
    \setcounter{enumi}{11}

\item I don’t fully understand why you consider “countable” many random variables instead of just a finite set of random variables. It looks to me that
all the results. I’m not an expert in probability theory but in this way,
you can represent random processes as for instance $X_i \sim N (\mu = X_i , \sigma)$,
and this might introduce rather complex aspects. Perhaps you should
comment on the case of infinite random variables.

\commentrebuttal{We are not entirely sure what the reviewer is referring to: first the reviewer talks about "countably many" (which can be an infinite number but still countable) then in the last sentence the reviewer mentioned infinite random variables. We are not sure whether the reviewer means  countable or uncountable infinity here.

Concerning the distribution above ($X_i \sim N (\mu = X_i , \sigma)$). Here we have a cycle: $X_i$ is used as its own mean and we do not have a valid random variable to begin with. 
}

\item Definition 3.10 Well-Defined Distributional Database. This coincides with
the definition of Bayesian network. I might have missed some details but
I don’t see the difference when the set of random variables is finite. When
they are infinite, the restriction that every formula has a finite number of
ancestors

\commentrebuttal{We substantially altered the definition of a distritbutional database (Definition~\ref{def:distDB}) and the reviewer's comment does directly apply anymore. We point, however, to the new Section~\ref{sec:fintiedistdb}, which we hope clarifies possible remaining doubts.
}

\item Condition DC1 This condition is semantically defined. How is it possible
to check it at the syntactic level without running the program itself? The
conditions might depend on the entire part of the program e.g., Suppose
that you have two statements
\begin{align*}
    x \sim \delta_1 \lpif A.
    \\
    x \sim \delta_2 \lpif B.
\end{align*}
2and there are two subpart of the program that imposes conditions on A
and B may depend on other random variables. As a matter of fact, you
have to ensure that $P(A \land B)$ is equal to $0$. I believe you have to provide
some sufficient “syntactic” condition that guarantees such that $A \land B$ is
not satisfiable.

\commentrebuttal{Indeed, when implementing the language one could think about introducing syntactic sugar that would guarantee validity.
The problem with such syntactic constructs is that they might be sufficient but not necessary to guarantee validity and thereby they would limit the user.

In our opinion ensuring that a set of distributional clauses is well-defined lies therefore with the user of the language itself or with the programmer who implements the language. In this paper we are, however, concerned with the semantics and therefore refrain from making such restrictions. 
}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/arithmetics_old_20210120.tex
aij dcproblog paper

\section{Typing \dcproblogsty}
\label{app:types}

\subsection{Type System}\label{sec:type_system}

A distinctive property of system predicates, which are implemented using an external engine, is their mandatory type signature.
An instance of such system predicates are comparison predicates.
Consider for example the comparison (\probloginline{x>20}).
The right-hand side is an integer and if the \probloginline{x} denotes a numeric random variable we have a well-typed comparison predicate inducing a measurable set.
Consider for the sake of contrast also the comparison predicate (\probloginline{a>2}), where \probloginline{a} is an atomic formula. This comparison predicate is ill-typed as atomic formulas cannot be compared to (integer) numbers. In other words, the external engine, which implements the system predicates does not know how to commensurate an atomic formula and a number.
A type system lets us specify which combinations of types are allowed as arguments of comparison predicates, \ie which comparisons encode measurable sets and are supported by the external engine.

\begin{samepage}
\begin{definition}\label{def:type_system}
	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes.
	The base type for \dcproblogsty is the \type{Term} type. We subtype terms with a \type{Distribution} type and a \type{Symbol} type. 
    A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.
	\begin{enumerate}
		\item[] \type{Term} 
		\begin{enumerate}
			\item[1.] \type{Distribution}
			\begin{itemize}
				\item \type{NumericD}
				\item \type{CategorialD}
			\end{itemize}
			\item[2.] \type{Symbol}
			\begin{itemize}		
				\item \type{RandomSymbol}
				\begin{itemize}
				    \item \type{NumericRV}
				    \item \type{CategoricalRV}
				\end{itemize}
				\item \type{DeterministicSymbol}
				\begin{itemize}
				    \item \type{Number}
				    \item \type{Atom}
				\end{itemize}	
			\end{itemize}
		\end{enumerate}
	\end{enumerate}
\end{definition}
\end{samepage}

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=\linewidth]{../figures/type_hierarchy_reduced.pdf}
	\end{center}
	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system.}
	\label{fig:type_system}
\end{figure}

\begin{definition}[\type{Distribution}]
	Terms of type \type{Distribution} are constructed by using a reserved random function symbol and applying it to a (possible empty) tuple of input arguments, which maps the input tuple to a distribution term.
	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{~}\lstinline[columns=fixed]|/2|.
	This also means that random function symbols are only used there as well. 
% 	\dcproblogsty subtypes the \type{Distribution} type into \type{NumericD} and \type{CategoricalD}.
% 	\type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
\end{definition}

\type{Distribution} terms determine the probability distribution according to which a random variable is distributed.
\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of type \type{NumericD}.

% Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}. \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
%\begin{example} \probloginline{mixed} is a random variables that is a mixture of a continuous part (the truncated normal between $0$ and $4$ with mean $3$ and standard deviation $1$) and a discrete part. 
%\begin{problog}
%mixted~Mix((0.998,TruncNormal(3,1,0,4)),(0.001,4),(0.001,4))
%\end{problog}
%\end{example}

@@

\begin{definition}[\type{Symbol}]
  \type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program. Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms (this includes standard \prologsty terms such as lists).  
\end{definition}

%The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.

@@

\begin{definition}[\type{RandomSymbol}] \label{def:random_variable}
	\type{RandomSymbol} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{~}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term. All random variables present in a \dcproblogsty program  explicitly named.
\end{definition}

Examples of terms of type \type{RandomSymbol} would be \probloginline{temperature} and \probloginline{n_people} in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, where we declared the random variables through distributional facts: 
\begin{problog}
temperature ~ normal(20,5).
n_people ~ poisson(6).
\end{problog}
As mentioned in Defintion~\ref{def:random_variable}, \type{RandomSymbol} terms inherit their type from the associated \type{Distribution} term, \eg a \type{NumericD} term induces a random variable term of type \type{NumericRV}.

Note that not all probabilistic programming languages enforce the naming of random variables. In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions. 
\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice} \pedro{usually in logic programming everything is global. an unamed random variable would mean that it is only available locally. we could allow that, I guess but would have to introduce new syntax. moreover, internally we would give the unnamed random variable a name anyway.}

\begin{definition}[\type{Atom}]
	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \dcplpinline{p} is a predicate with arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms. Functors used to construct \type{Atom} terms must not clash with the builtin predicates and reserved functors used to declare \type{Distribution} terms.
% 	Terms of type \type{Atom} constitute the building blocks of the discrete structure of a probabilistic logic program.
\end{definition}

Examples of \type{Atom} terms would be \probloginline{foo} or \probloginline{bar(foo,42)}. The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}. \probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.

@@

While \prologsty is traditionally not strongly/statically typed, there have been efforts to extend \prologsty with such a type system. Arguably, the most interesting effort is presented in~\citep{Schrijvers2008TowardsTP}, where the authors develop the idea of partially typing a \prologsty program, \ie typing is optional and untyped code is interpreted instead of compiled. Introducing predicates with strongly typed signatures can be regarded as adding predicates with a specific meaning to the \prologsty implementation but the {\em `evaluation'} of these predicates still takes place within \prologsty.
Our efforts of introducing a type system for system predicates within a probabilistic programming context are orthogonal and complimentary to the efforts of~\citet{Schrijvers2008TowardsTP}.

@@

\subsection{External Engine}\label{sec:external_engine}

In the previous subsection we have introduced an abstract type system.
Abstract types describe the structure/hierarchy of a type system and cannot be instantiated. Concrete types, which are subtypes of abstract types are used for instantiations. This means that abstract types relate collections of concrete types, \ie implementations, to each other. For example, the \type{Integer} \probloginline{3} could be instantiated as an \type{Int32} or also an \type{Int64}, which are both concrete types. The choice taken depends on the external engine used to evaluate arithmetic expressions.

The only condition that we necessitate for an implementation of \dcproblogsty is that the external engine provides concrete types for the abstract types. Additionally, a practical external engine should also implement operations on these types. For example, an engine should be able to multiply a floating point value and an integer value: (\probloginline{3.12*2}). The type for floating points could be dubbed \type{Float32} and would be a concrete type subtyped from the abstract type \type{Real}. The type of the integer value could be \type{Int32} subtyped from \type{Integer}. The return type could then again be \type{Float32}.

\begin{example}\label{example:julia_addition}
We show two functions that perform addition of numbers but with different types. These different number could be invoked on either side of a comparison predicate, \eg (\probloginline{1<3+5} ) or (\probloginline{1<3+5.0} ). We write these functions using Julia like syntax~\citep{bezanson2017julia}. 
\begin{samepage}

>>>>

%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%

\section{Arithmetic Expressions}\label{sec:arithmetic}
\ak{I think this should be a part of implementation section, and should be about comparison predicates rather than about arithmetic expressions}

Implementations of (probabilistic) logic programming languages, such as \prologsty and \problogsty, reserve a set of predicate names for system-related procedures. These system predicates (or built-in predicates) are usually deployed when the goal is to perform efficient arithmetic operations on numbers. Either by using specialized algorithms or specialized hardware. In \dcproblogsty we are going to use an external arithmetic engine to evaluate expressions within comparison predicates.

\begin{example}\label{ex:eval_comp} A \dcproblogsty code snippet demonstrating arithmetic evaluations within a comparison predicate.\ak{how is this example relevant here?}
	\begin{problog*}{linenos}
0.5::a.
x ~ normal(0,3):- a.
x ~ normal(1,2):- \+a.

q:- x+3>3.
query(q).
	\end{problog*}
Using the \probloginline{map}\lstinline[columns=fixed]|/2| predicate this program gets transformed into the program below.
	\begin{problog*}{linenos}
0.5::a.
x1 ~ normal(0,3).
map(x1,x):- a.

x2 ~ normal(1,2).
map(x2,x):- \+a.

q:- map(X,x), X+3>3.
query(q).
	\end{problog*}

\end{example}

\luc{technical problem with DC; do we need this?}

\subsection{Type System} \label{sec:type_system}

While we have stated that arithmetic expressions can appear on both sides of comparison predicates, we have not yet mentioned what they evaluate to.
To this end we will regard expressions to be evaluated as unnamed functions, instead of arithmetic predicates\ak{we do not have arithmetic predicates, only arithmetic functors}. Let us consider, for instance, the expression \probloginline{3+5}. This can can be viewed as a term whose predicate is \probloginline{+}\lstinline[columns=fixed]|/2| in infix notation. However, we can also regard the expression as a function with two inputs and an output. Furthermore, we can equip the function with a type signature. The input values are of type \type{Integer} and the output value is of \type{Integer} as well. We will now introduce a type system for the evaluation of arithmetic expressions in \dcproblogsty.

The base type for (arithmetic evaluations in) \dcproblogsty is the \type{Term} type. We subtype terms with a \type{Distribution} type and a \type{Symbol} type. 
Dividing \dcproblogsty terms into \justify\type{Distribution} terms and \type{Symbol} terms reflects the characterization of~\citet{poole2010probabilistic} of probabilistic programs as being constituted of independent choices (\type{Distribution}) and a deterministic system (\type{Symbol}).

\begin{definition}\label{def:type_system}
	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes. A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.
	\begin{enumerate}
		\item[] \type{Term} 
		\begin{enumerate}
			\item[1.] \type{Distribution}
			\begin{itemize}
				\item \type{NumericD}
				\item \type{CategorialD}
			\end{itemize}
			\item[2.] \type{Symbol}
			\begin{itemize}		
				\item \type{RandomSymbol}
				\begin{itemize}
				    \item \type{NumericRV}
				    \item \type{CategoricalRV}
				\end{itemize}
				\item \type{DeterministicSymbol}
				\begin{itemize}
				    \item \type{Number}
				    \item \type{Atom}
				\end{itemize}	
			\end{itemize}
		\end{enumerate}
	\end{enumerate}
\end{definition}

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=\linewidth]{../figures/type_hierarchy.pdf}
	\end{center}
	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system. Compared to Definition~\ref{def:type_system},  the \type{Number} type is further subtyped into \type{Integer} and \type{Real} types and the \type{RandomVariable} type is subtyped into \type{DiscreteRV}, \type{ContinuousRV}, and \type{MixtureRV}. These subtypes of \type{RandomVariable} mimic the subtypes of the \type{Distribution} type.\luc{do we want mixtureRV?}}
	\label{fig:type_system}
\end{figure}

\begin{definition}[\type{Distribution}]
	Terms of type \type{Distribution} are constructed by using a reserved random function symbol and applying it to a (possible empty) tuple of input arguments, which maps the input tuple to a distribution term.
	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{~}\lstinline[columns=fixed]|/2|.
	This also means that random function symbols are only used there as well. \dcproblogsty subtypes the \type{Distribution} type into \type{NumericD} and \type{CategoricalD}.
% 	\type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
\end{definition}

\type{Distribution} terms determine the probability distribution according to which a random variable is distributed.
\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of types \type{DiscreteD} and  \type{ContinuousD}, respectively.
Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}. \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
%\begin{example} \probloginline{mixed} is a random variables that is a mixture of a continuous part (the truncated normal between $0$ and $4$ with mean $3$ and standard deviation $1$) and a discrete part. 
%\begin{problog}
%mixted~Mix((0.998,TruncNormal(3,1,0,4)),(0.001,4),(0.001,4))
%\end{problog}
%\end{example}

@@

\type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program. Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms -- this includes standard \prologsty terms such as lists.
%The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.

@@

\begin{definition}[\type{RandomSymbol}] \label{def:random_variable}
	\type{RandomSymbol} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{~}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term. All random variables present in a \dcproblogsty program are named.
\end{definition}

Examples of terms of type \type{RandomSymbol} would be \probloginline{temperature} and \probloginline{n_people} in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, where we declared the random variables through distributional facts: 
\begin{problog}
temperature ~ normal(20,5).
n_people ~ poisson(6).
\end{problog}
As mentioned in Defintion~\ref{def:random_variable}, \type{RandomSymbol} terms inherit their type from the associated \type{Distribution} term, \eg a \type{ContinuousD} term induces a random variable term of type \type{ContinuousRV}.

Note that not all probabilistic programming languages enforce the naming of random variables. In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions.
\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice}

@@

\ak{terminology mix-up; predicates never form terms} \pedro{not sure what this comment refers to.}

\begin{definition}[\type{Atom}]
	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \probloginline{p} is a predicate of arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms~\citep{fierens2015inference}. Predicates used to construct \type{Atom} terms must not clash with the builtin predicates and reserved random function symbols used to declare \type{Distribution} terms.
% 	Terms of type \type{Atom} constitute the building blocks of the discrete structure of a probabilistic logic program.
\end{definition}

Examples of atoms terms would be \probloginline{foo} or \probloginline{bar(foo,42)}. The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}. \probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.

@@

\subsection{Multiple Dispatch} \label{ref:multiple_dispatch}
\ak{I think there's a lot in here that misses context to be understandable; and I'm not sure all of it is really needed for the paper}
The attentive reader might have noticed that, for example in Example~\ref{ex:dc-vs-dcproblog}, we used the same comparison predicates, for \type{RandomSymbol} terms as well as \type{Number} terms. In \dcproblogsty we resolve this clash through {\em multiple dispatching}~\citep{castagna1995calculus}.
Multiple dispatching is the process of dynamically selecting at run time\footnote{In the context of logic programming we understand as run time the grounding of the logic program part of a \dcproblogsty program.} a specific implementation of a function based on the types of the arguments that are fed as input to the corresponding function symbol.
Such functions are referred to as {\em multimethods}. 
An early example of multiple dispatch is the {\em Common Lisp Object System}~\citep{keene1989object}. Multiple dispatch is also the programming paradigm on which the \juliasty programming language is built~\citep{bezanson2017julia}.

We also find multimethods in logic programming languages such as \prologsty. While, for groundings of relations \prologsty simply checks whether the grounding is present in the database (agnostic towards the types of the arguments), this is not the case for built-in predicates such as comparison predicates, which make a call to an external arithmetic engine in order to evaluate an expression. For instance, (\probloginline{1<3+5}) calls an integer addition procedure, whereas (\probloginline{1<3.0+5}) might call a floating point procedure. Calling the external arithmetic engine with an ill-typed expression might even cause a run time error. Something that is not possible in {\em pure} \prologsty without an external arithmetic engine.
